---
title: "Kev-Carbon: Bayesian demographic transitions in Australian radiocarbon data"
author: "Kevan Edinborough"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: readable
params:
  run_mode: "test"
  start_calbp: 2000          # older boundary; 2000 BP is approximately 50 BC
  end_calbp: 0               # radiocarbon present = AD 1950
  calibration_curve: "shcal20"
  calendar_display: "calAD"   # "calAD" or "calBP"
  parallel_cores: 1          # forced to 1 on this 8 GB laptop
  mcmc_chains: null           # null uses run_mode preset
  mcmc_iterations: null       # total iterations per chain
  mcmc_burnin: null           # discarded iterations per chain
  mcmc_thin: null             # retain every nth iteration
  mcmc_seed: 73021
---

# Aim

This vignette compares three Bayesian demographic models for the same Australian radiocarbon dataset over a user-defined calendar interval. The default interval is 2000–0 cal BP (approximately 50 BC–AD 1950), providing a substantial pre-colonial baseline as well as the colonial period:

1. **Single-slope model:** one exponential rate across the complete interval. In plain English, the radiocarbon-derived trajectory changes gradually at approximately one proportional rate; the data do not require a distinct break around colonisation.
2. **One-change-point model:** one rate before and another after an estimated change point.
3. **Two-change-point model:** three successive demographic rates separated by two estimated transition dates. In plain English, this allows the evidence to show growth, decline and recovery — or other three-phase trajectories — without fixing the transition dates in advance.

The radiocarbon dates are calibrated with `rcarbon`. Site-time bins are used to reduce oversampling, and one median calibrated age is retained per bin. Two models use distributions supplied by `nimbleCarbon`; the third is a Kev-Carbon extension with two change points. All three are compared using WAIC.

> **Run note:** the two-change-point model is a new custom NIMBLE distribution. Always run `test` mode first. A successful test means the model compiles, samples, writes the output tables and produces all five graphics before you commit to a medium or long run.

> **Interpretive limit:** this is a model of the temporal distribution of archaeological radiocarbon evidence. It does not prove that European colonisation caused a demographic decline. A colonial interpretation requires the estimated change point and its uncertainty to be compared with independent historical evidence, while considering sampling, preservation, research intensity, land-use change, mobility and altered depositional practices.

> **McDead laptop mode:** this version is configured for an Intel i5-8250U with 8 GB RAM. It never launches parallel NIMBLE workers, fits the two models sequentially, uses smaller run presets, disables caching of compiled model fits, and limits posterior curve draws to reduce memory pressure.

# 1. Setup

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.width = 9,
  fig.height = 5.5
)

required <- c(
  "readxl", "dplyr", "ggplot2", "tibble", "tidyr",
  "rcarbon", "nimble", "nimbleCarbon", "coda", "knitr", "pkgbuild"
)

missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing)) {
  stop(
    "Install the missing packages before knitting: ",
    paste(missing, collapse = ", "),
    "\nRun: install.packages(c(",
    paste(sprintf('"%s"', missing), collapse = ", "),
    "))"
  )
}

library(readxl)
library(dplyr)
library(ggplot2)
library(tibble)
library(tidyr)
library(rcarbon)
library(nimble)
library(nimbleCarbon)
library(coda)
library(knitr)

dir.create("output", showWarnings = FALSE, recursive = TRUE)


# -------------------------------------------------------------------------
# WINDOWS COMPILER PREFLIGHT
# NIMBLE generates and compiles C++ code. On Windows this requires the
# Rtools version matching the installed major/minor R version.
# -------------------------------------------------------------------------
if (.Platform$OS.type == "windows") {
  r_minor <- paste(R.version$major, strsplit(R.version$minor, "\\.")[[1]][1], sep = ".")
  expected_rtools <- switch(
    r_minor,
    "4.5" = "Rtools45",
    "4.4" = "Rtools44",
    "4.3" = "Rtools43",
    paste0("the Rtools release matching R ", r_minor)
  )

  build_ok <- tryCatch(
    pkgbuild::has_build_tools(debug = TRUE),
    error = function(e) FALSE
  )

  make_path <- Sys.which("make")
  gpp_path  <- Sys.which("g++")

  compiler_check <- tibble::tibble(
    item = c("R version", "Required Windows toolchain", "Build tools detected",
             "make", "g++", "Temporary directory", "Free-memory guidance"),
    value = c(
      R.version.string,
      expected_rtools,
      build_ok,
      ifelse(nzchar(make_path), make_path, "NOT FOUND"),
      ifelse(nzchar(gpp_path), gpp_path, "NOT FOUND"),
      tempdir(),
      "Close browsers and other large programs before fitting"
    )
  )
  print(knitr::kable(compiler_check))

  if (!isTRUE(build_ok) || !nzchar(make_path) || !nzchar(gpp_path)) {
    stop(
      paste0(
        "NIMBLE cannot compile because the Windows C++ build tools are unavailable.\\n\\n",
        "Installed R: ", R.version.string, "\\n",
        "Required toolchain: ", expected_rtools, "\\n\\n",
        "Fix:\\n",
        "1. Close RStudio.\\n",
        "2. Install ", expected_rtools, " from CRAN using the default location and options.\\n",
        "3. Restart Windows, then reopen RStudio.\\n",
        "4. Run pkgbuild::check_build_tools(debug = TRUE).\\n",
        "5. Continue only when it reports that the system is ready to build packages.\\n\\n",
        "Detected make: ", ifelse(nzchar(make_path), make_path, "NOT FOUND"), "\\n",
        "Detected g++: ", ifelse(nzchar(gpp_path), gpp_path, "NOT FOUND")
      ),
      call. = FALSE
    )
  }
}

# -------------------------------------------------------------------------
# GLOBAL PARAMETERS
# Edit these in the YAML header at the top of this vignette.
# calendar_display = "calAD" or "calBP" controls tables and figure axes.
# parallel_cores = 1 runs sequentially; 2 runs the two model fits in parallel.
# Because this vignette compares only two models, values above 2 are capped at 2.
# Explicit MCMC values override the selected run_mode preset. Leave them null
# to use the test, medium, or long defaults.
# -------------------------------------------------------------------------

# Conservative presets for McDead: i5-8250U, 8 GB RAM and limited free disk.
# Models and chains run sequentially to avoid simultaneous C++ compilation and
# the shared-library failures produced by Windows PSOCK workers.
run_presets <- list(
  test   = list(niter = 2000,  nburnin = 700,   thin = 5,  nchains = 1),
  medium = list(niter = 12000, nburnin = 4000,  thin = 8,  nchains = 2),
  long   = list(niter = 40000, nburnin = 15000, thin = 15, nchains = 2)
)

run_mode <- tolower(params$run_mode)
if (!run_mode %in% names(run_presets)) {
  stop("run_mode must be 'test', 'medium', or 'long'.")
}

mcmc_cfg <- run_presets[[run_mode]]

resolve_positive_integer <- function(value, default, label, allow_zero = FALSE) {
  if (is.null(value) || length(value) == 0 || is.na(value)) return(as.integer(default))
  out <- suppressWarnings(as.integer(value))
  minimum <- if (allow_zero) 0L else 1L
  if (is.na(out) || out < minimum || out != as.numeric(value)) {
    stop(label, " must be a whole number of ", minimum, " or more, or null.")
  }
  out
}

mcmc_cfg$nchains <- resolve_positive_integer(params$mcmc_chains, mcmc_cfg$nchains, "mcmc_chains")
mcmc_cfg$niter <- resolve_positive_integer(params$mcmc_iterations, mcmc_cfg$niter, "mcmc_iterations")
mcmc_cfg$nburnin <- resolve_positive_integer(params$mcmc_burnin, mcmc_cfg$nburnin, "mcmc_burnin", allow_zero = TRUE)
mcmc_cfg$thin <- resolve_positive_integer(params$mcmc_thin, mcmc_cfg$thin, "mcmc_thin")
mcmc_cfg$seed <- resolve_positive_integer(params$mcmc_seed, 73021L, "mcmc_seed")

if (mcmc_cfg$nburnin >= mcmc_cfg$niter) {
  stop("mcmc_burnin must be smaller than mcmc_iterations.")
}
if ((mcmc_cfg$niter - mcmc_cfg$nburnin) %/% mcmc_cfg$thin < 50) {
  warning("Fewer than 50 posterior samples per chain will be retained. Increase iterations or reduce thinning.")
}

start_bp <- as.numeric(params$start_calbp)
end_bp <- as.numeric(params$end_calbp)
calendar_display <- tolower(as.character(params$calendar_display))
if (!calendar_display %in% c("calad", "calbp")) {
  stop("calendar_display must be 'calAD' or 'calBP'.")
}

bp_to_ad <- function(x) 1950 - x
calendar_axis_label <- if (calendar_display == "calad") "Calendar year AD" else "Calendar years BP"
calendar_values <- function(x) if (calendar_display == "calad") bp_to_ad(x) else x
calendar_scale <- if (calendar_display == "calbp") {
  ggplot2::scale_x_reverse(limits = c(start_bp, end_bp))
} else {
  ggplot2::scale_x_continuous(limits = sort(c(bp_to_ad(start_bp), bp_to_ad(end_bp))))
}

available_cores <- parallel::detectCores(logical = TRUE)
if (is.na(available_cores) || available_cores < 1) available_cores <- 1L
requested_cores <- suppressWarnings(as.integer(params$parallel_cores))
if (is.na(requested_cores) || requested_cores < 1L) {
  stop("parallel_cores must be a whole number of 1 or more.")
}

# HARD SAFETY LIMIT FOR THIS MACHINE.
# NIMBLE compilation is deliberately sequential. The i5-8250U has four cores,
# but only 8 GB RAM; two Windows workers compiling DLLs concurrently caused the
# checkForRemoteErrors()/shared-library failure seen during knitting.
parallel_cores <- 1L
parallel_enabled <- FALSE
if (requested_cores > 1L) {
  warning("parallel_cores has been forced to 1 for McDead (8 GB RAM).")
}

# Fail early if the working drive is extremely full. NIMBLE and R need room for
# generated C++, object files, DLLs, caches and temporary report files.
free_space_note <- "Keep at least 10 GB free on the Windows drive before a medium or long run."
cat("\n**Machine safety:** sequential NIMBLE compilation; ", free_space_note, "\n", sep = "")

if (!is.finite(start_bp) || !is.finite(end_bp) || start_bp <= end_bp) {
  stop("start_calbp must be older/larger than end_calbp.")
}

set.seed(mcmc_cfg$seed)

configuration <- tibble(
  setting = c("Run mode", "Analysis start cal BP", "Analysis end cal BP",
              "Displayed calendar scale", "Displayed start", "Displayed end",
              "Calibration curve", "Iterations per chain", "Burn-in", "Thinning",
              "Chains", "MCMC seed", "Retained samples per chain",
              "Logical processors detected", "Processors requested",
              "Processors used", "Parallel model fitting"),
  value = c(run_mode, start_bp, end_bp, ifelse(calendar_display == "calad", "cal AD", "cal BP"),
            calendar_values(start_bp), calendar_values(end_bp), params$calibration_curve,
            mcmc_cfg$niter, mcmc_cfg$nburnin, mcmc_cfg$thin, mcmc_cfg$nchains,
            mcmc_cfg$seed, (mcmc_cfg$niter - mcmc_cfg$nburnin) %/% mcmc_cfg$thin,
            available_cores, requested_cores, parallel_cores, parallel_enabled)
)
knitr::kable(configuration)

window_years <- start_bp - end_bp
window_start_ad <- bp_to_ad(start_bp)
window_end_ad <- bp_to_ad(end_bp)

cat("\n**Resolved model window:** ", start_bp, "–", end_bp, " cal BP (approximately ",
    window_start_ad, " to ", window_end_ad, " AD; ", window_years, " years).\n", sep = "")

if (window_years < 800) {
  warning("The model window is shorter than 800 years. It may not provide enough pre-colonial baseline to identify the full curve or a stable change point.")
}
if (window_end_ad < 1788 || window_start_ad > 1850) {
  warning("The selected interval does not fully bracket the principal period of European colonisation. Review start_calbp and end_calbp.")
}
```

# 2. Import the same AusData dataset

```{r import}
input_file <- "AusData.xlsx"
if (!file.exists(input_file)) {
  stop("AusData.xlsx is not in the project folder.")
}

raw <- readxl::read_excel(input_file, sheet = "AusData")

needed <- c("LabID", "SiteID", "C14Age", "C14Error")
missing_cols <- setdiff(needed, names(raw))
if (length(missing_cols)) {
  stop("Missing required columns: ", paste(missing_cols, collapse = ", "))
}

AUD <- raw |>
  transmute(
    lab_id = as.character(LabID),
    site_id = as.character(SiteID),
    c14_age = suppressWarnings(as.numeric(C14Age)),
    c14_error = suppressWarnings(as.numeric(C14Error))
  ) |>
  filter(
    !is.na(lab_id),
    !is.na(site_id),
    is.finite(c14_age),
    is.finite(c14_error),
    c14_error > 0
  ) |>
  mutate(
    # rcarbon requires every calibration ID to be unique. Laboratory IDs are
    # retained separately because replicate determinations can share a LabID.
    calibration_id = sprintf("AusDate_%05d", row_number())
  )

audit <- tibble(
  rows_in_workbook = nrow(raw),
  valid_dates = nrow(AUD),
  sites = n_distinct(AUD$site_id),
  median_error = median(AUD$c14_error)
)
knitr::kable(audit, digits = 1)
```

# 3. Calibrate and create independent site-time bins

```{r calibrate-bin}
cal <- rcarbon::calibrate(
  x = AUD$c14_age,
  errors = AUD$c14_error,
  ids = AUD$calibration_id,
  calCurves = params$calibration_curve,
  normalised = FALSE,
  verbose = TRUE
)

cal_summary <- as.data.frame(summary(cal))
if (!"MedianBP" %in% names(cal_summary)) {
  stop("The rcarbon calibration summary did not contain MedianBP.")
}

AUD$median_calbp <- suppressWarnings(as.numeric(cal_summary$MedianBP))

# Group same-site determinations that are close in conventional radiocarbon age.
AUD$bin_id <- rcarbon::binPrep(
  sites = AUD$site_id,
  ages = AUD$c14_age,
  h = 100
)

# One representative calendar estimate per site-time bin.
binned <- AUD |>
  filter(
    is.finite(median_calbp),
    median_calbp >= end_bp,
    median_calbp <= start_bp
  ) |>
  group_by(bin_id) |>
  summarise(
    theta = median(median_calbp),
    site_id = first(site_id),
    dates_in_bin = n(),
    .groups = "drop"
  ) |>
  arrange(desc(theta))

if (nrow(binned) < 20) {
  stop("Fewer than 20 site-time bins fall inside the selected analysis window.")
}

cat("Independent site-time bins used:", nrow(binned), "\n")
cat("Original determinations represented:", sum(binned$dates_in_bin), "\n")

write.csv(binned, "output/binned_calendar_dates.csv", row.names = FALSE)

binned_plot <- binned |> mutate(calendar_display_value = calendar_values(theta))

 ggplot(binned_plot, aes(calendar_display_value)) +
  geom_histogram(binwidth = 25, boundary = 0) +
  calendar_scale +
  labs(
    title = "Binned calibrated dates used by both models",
    x = calendar_axis_label,
    y = "Site-time bins"
  ) +
  theme_minimal(base_size = 12)
```


# 4. The three demographic models

The purpose of this section is simple: ask whether the radiocarbon evidence is best described by **one broad trend**, **one major change**, or **two major changes**.

**Model 1 — single slope.** One demographic rate is estimated across the complete interval. This is the simplest explanation.

**Model 2 — one change point.** An early rate and a late rate are separated by one estimated transition date. This is already available in `nimbleCarbon` as `dDoubleExponentialGrowth()`.

**Model 3 — two change points (Kev-Carbon extension).** Three rates are separated by two estimated transition dates. This allows trajectories such as growth → decline → recovery, or stability → decline → steeper decline.

> **Plain English:** the new model does not assume that Australian demographic change happened at a historically chosen date. It lets the radiocarbon evidence estimate two transition dates and quantifies the uncertainty around them.

## 4.1 Kev-Carbon custom demographic distribution

The custom distribution below follows the same annual-grid logic as the demographic distributions used by `nimbleCarbon`. The three pieces are joined continuously, converted to positive relative demographic intensity, and normalised to a probability distribution over calendar time.

```{r kevcarbon-distribution}
dTripleExponentialGrowth <- nimble::nimbleFunction(
  run = function(
    x   = double(0),
    a   = double(0),
    b   = double(0),
    r1  = double(0),
    r2  = double(0),
    r3  = double(0),
    mu1 = double(0),
    mu2 = double(0),
    log = integer(0, default = 0)
  ) {
    returnType(double(0))

    mu1r <- round(mu1)
    mu2r <- round(mu2)

    # Reject impossible change-point ordering.
    if (mu1r >= a || mu2r >= mu1r || mu2r <= b) {
      if (log == 1) return(-Inf)
      return(0.0)
    }

    n <- round(a - b + 1)
    weights <- numeric(n)

    # Index runs from the oldest boundary a toward the present boundary b.
    for (i in 1:n) {
      yr <- a - (i - 1)

      if (yr > mu1r) {
        elapsed1 <- a - yr
        weights[i] <- exp(r1 * elapsed1)
      } else if (yr > mu2r) {
        elapsed1 <- a - mu1r
        elapsed2 <- mu1r - yr
        weights[i] <- exp(r1 * elapsed1 + r2 * elapsed2)
      } else {
        elapsed1 <- a - mu1r
        elapsed2 <- mu1r - mu2r
        elapsed3 <- mu2r - yr
        weights[i] <- exp(
          r1 * elapsed1 +
          r2 * elapsed2 +
          r3 * elapsed3
        )
      }
    }

    total <- sum(weights)
    idx <- round(a - x + 1)

    if (idx < 1 || idx > n || total <= 0) {
      if (log == 1) return(-Inf)
      return(0.0)
    }

    p <- weights[idx] / total

    if (log == 1) return(log(p))
    return(p)
  }
)

rTripleExponentialGrowth <- nimble::nimbleFunction(
  run = function(
    n   = integer(0),
    a   = double(0),
    b   = double(0),
    r1  = double(0),
    r2  = double(0),
    r3  = double(0),
    mu1 = double(0),
    mu2 = double(0)
  ) {
    returnType(double(0))

    mu1r <- round(mu1)
    mu2r <- round(mu2)
    ny <- round(a - b + 1)
    weights <- numeric(ny)

    for (i in 1:ny) {
      yr <- a - (i - 1)

      if (yr > mu1r) {
        elapsed1 <- a - yr
        weights[i] <- exp(r1 * elapsed1)
      } else if (yr > mu2r) {
        elapsed1 <- a - mu1r
        elapsed2 <- mu1r - yr
        weights[i] <- exp(r1 * elapsed1 + r2 * elapsed2)
      } else {
        elapsed1 <- a - mu1r
        elapsed2 <- mu1r - mu2r
        elapsed3 <- mu2r - yr
        weights[i] <- exp(
          r1 * elapsed1 +
          r2 * elapsed2 +
          r3 * elapsed3
        )
      }
    }

    probs <- weights / sum(weights)
    draw <- rcat(1, probs)
    return(a - draw + 1)
  }
)

# Vectorised R helper used only for plotting.
triple_density_vec <- function(x, a, b, r1, r2, r3, mu1, mu2) {
  elapsed1 <- pmax(0, pmin(a - x, a - mu1))
  elapsed2 <- pmax(0, pmin(mu1 - x, mu1 - mu2))
  elapsed3 <- pmax(0, mu2 - x)

  log_w <- r1 * elapsed1 + r2 * elapsed2 + r3 * elapsed3
  w <- exp(log_w - max(log_w))
  w / sum(w)
}
```

## 4.2 NIMBLE model code

```{r model-code}
theta <- binned$theta
N <- length(theta)

single_slope_code <- nimble::nimbleCode({
  r ~ dnorm(0, sd = 0.02)
  for (i in 1:N) {
    theta[i] ~ dExponentialGrowth(a = startBP, b = endBP, r = r)
  }
})

change_point_code <- nimble::nimbleCode({
  r_early ~ dnorm(0, sd = 0.02)
  r_late  ~ dnorm(0, sd = 0.02)
  mu ~ dunif(changeLow, changeHigh)

  for (i in 1:N) {
    theta[i] ~ dDoubleExponentialGrowth(
      a = startBP, b = endBP,
      r1 = r_early, r2 = r_late, mu = mu
    )
  }
})

triple_change_code <- nimble::nimbleCode({
  r1 ~ dnorm(0, sd = 0.02)
  r2 ~ dnorm(0, sd = 0.02)
  r3 ~ dnorm(0, sd = 0.02)

  # mu1 is the older transition.
  mu1 ~ dunif(change2Low, change1High)

  # The second transition is forced to be younger than mu1.
  separation ~ dunif(minSeparation, maxSeparation)
  mu2 <- mu1 - separation

  for (i in 1:N) {
    theta[i] ~ dTripleExponentialGrowth(
      a = startBP, b = endBP,
      r1 = r1, r2 = r2, r3 = r3,
      mu1 = mu1, mu2 = mu2
    )
  }
})
```

# 5. Fit all three models

The test mode is deliberately small. Use it only to establish that compilation, sampling and graphics work. Medium and long runs are for inference.

```{r fit-models, cache=FALSE}
extract_numeric_waic <- function(x, depth = 0L) {
  if (is.null(x) || depth > 8L) return(NA_real_)

  if (is.numeric(x)) {
    out <- suppressWarnings(as.numeric(x))
    out <- out[is.finite(out)]
    return(if (length(out)) out[1] else NA_real_)
  }

  candidate_names <- c(
    "WAIC", "waic", "value", "totalWAIC", "WAICvalue",
    "waicValue", "WAIC_total", "total"
  )

  if (is.list(x)) {
    for (nm in intersect(candidate_names, names(x))) {
      out <- extract_numeric_waic(x[[nm]], depth + 1L)
      if (is.finite(out)) return(out)
    }
    for (element in x) {
      out <- extract_numeric_waic(element, depth + 1L)
      if (is.finite(out)) return(out)
    }
  }

  if (isS4(x)) {
    slots <- methods::slotNames(x)
    for (nm in intersect(candidate_names, slots)) {
      out <- extract_numeric_waic(methods::slot(x, nm), depth + 1L)
      if (is.finite(out)) return(out)
    }
  }

  out <- tryCatch(suppressWarnings(as.numeric(x)), error = function(e) numeric())
  out <- out[is.finite(out)]
  if (length(out)) out[1] else NA_real_
}

extract_fit <- function(x) {
  samples <- if (is.list(x) && !is.null(x$samples)) x$samples else x
  waic_obj <- if (is.list(x) && !is.null(x$WAIC)) x$WAIC else attr(x, "WAIC", exact = TRUE)

  list(
    samples = samples,
    WAIC = extract_numeric_waic(waic_obj),
    WAIC_raw = waic_obj
  )
}

nimble_compiler_smoke_test <- function() {
  tiny_code <- nimble::nimbleCode({ x ~ dnorm(0, sd = 1) })
  tiny_model <- nimble::nimbleModel(
    code = tiny_code,
    inits = list(x = 0),
    check = TRUE,
    calculate = FALSE
  )

  tryCatch({
    tiny_compiled <- nimble::compileNimble(
      tiny_model,
      showCompilerOutput = TRUE,
      resetFunctions = TRUE
    )
    rm(tiny_compiled, tiny_model)
    invisible(gc())
    TRUE
  }, error = function(e) {
    try(nimble::printErrors(), silent = TRUE)
    stop(
      paste0(
        "NIMBLE compiler smoke test failed. Check the Rtools message above. ",
        conditionMessage(e)
      ),
      call. = FALSE
    )
  })
}

nimble_compiler_smoke_test()

fit_one_model <- function(model_name, theta, start_bp, end_bp, mcmc_cfg) {

  N <- length(theta)
  constants_common <- list(N = N, startBP = start_bp, endBP = end_bp)
  data_common <- list(theta = theta)

  if (model_name == "single") {
    code <- single_slope_code
    constants <- constants_common
    monitors <- "r"
    inits <- lapply(seq_len(mcmc_cfg$nchains), function(ch) {
      list(r = rnorm(1, 0, 0.002))
    })
  }

  if (model_name == "change") {
    code <- change_point_code
    constants <- c(
      constants_common,
      list(
        changeLow = end_bp + 100,
        changeHigh = start_bp - 100
      )
    )
    monitors <- c("r_early", "r_late", "mu")
    inits <- lapply(seq_len(mcmc_cfg$nchains), function(ch) {
      list(
        r_early = rnorm(1, 0, 0.002),
        r_late = rnorm(1, 0, 0.002),
        mu = runif(1, end_bp + 250, start_bp - 250)
      )
    })
  }

  if (model_name == "triple") {
    code <- triple_change_code

    usable_span <- start_bp - end_bp
    min_sep <- max(50, round(usable_span * 0.05))
    max_sep <- max(min_sep + 10, round(usable_span * 0.60))

    constants <- c(
      constants_common,
      list(
        change2Low = end_bp + min_sep + 50,
        change1High = start_bp - 50,
        minSeparation = min_sep,
        maxSeparation = max_sep
      )
    )

    monitors <- c("r1", "r2", "r3", "mu1", "mu2", "separation")

    inits <- lapply(seq_len(mcmc_cfg$nchains), function(ch) {
      set.seed(mcmc_cfg$seed + 300L + ch)

      mu1_init <- end_bp + 0.65 * usable_span
      sep_init <- min(max_sep - 5, max(min_sep + 5, 0.30 * usable_span))

      list(
        r1 = rnorm(1, 0, 0.002),
        r2 = rnorm(1, -0.001, 0.002),
        r3 = rnorm(1, 0, 0.002),
        mu1 = mu1_init,
        separation = sep_init
      )
    })
  }

  if (!exists("code")) stop("Unknown model: ", model_name)

  nimble::nimbleMCMC(
    code = code,
    constants = constants,
    data = data_common,
    inits = inits,
    monitors = monitors,
    niter = mcmc_cfg$niter,
    nburnin = mcmc_cfg$nburnin,
    thin = mcmc_cfg$thin,
    nchains = mcmc_cfg$nchains,
    setSeed = mcmc_cfg$seed +
      switch(model_name, single = 1000L, change = 2000L, triple = 3000L) +
      seq_len(mcmc_cfg$nchains),
    samples = TRUE,
    samplesAsCodaMCMC = TRUE,
    summary = FALSE,
    WAIC = TRUE,
    progressBar = FALSE
  )
}

fit_with_diagnostics <- function(model_name) {
  cat("\nStarting model:", model_name, "\n")

  tryCatch(
    fit_one_model(
      model_name,
      theta = theta,
      start_bp = start_bp,
      end_bp = end_bp,
      mcmc_cfg = mcmc_cfg
    ),
    error = function(e) {
      cat("\n========== NIMBLE ERROR:", toupper(model_name), "==========\n")
      try(nimble::printErrors(), silent = TRUE)
      stop(
        paste0("Model '", model_name, "' failed: ", conditionMessage(e)),
        call. = FALSE
      )
    }
  )
}

cat("Fitting three models sequentially to avoid simultaneous Windows compilation.\n")

single_raw <- fit_with_diagnostics("single")
gc(verbose = FALSE)

change_raw <- fit_with_diagnostics("change")
gc(verbose = FALSE)

triple_raw <- fit_with_diagnostics("triple")
gc(verbose = FALSE)

fit_single <- extract_fit(single_raw)
fit_change <- extract_fit(change_raw)
fit_triple <- extract_fit(triple_raw)

saveRDS(fit_single, "output/fit_single_slope.rds")
saveRDS(fit_change, "output/fit_change_point.rds")
saveRDS(fit_triple, "output/fit_two_change_points.rds")

cat("\nWAIC values\n")
cat("Single slope:       ", fit_single$WAIC, "\n")
cat("One change point:   ", fit_change$WAIC, "\n")
cat("Two change points:  ", fit_triple$WAIC, "\n")
```

# 6. Model comparison — which explanation wins?

```{r model-comparison}
comparison <- tibble(
  model = c(
    "Single slope",
    "One change point",
    "Two change points (Kev-Carbon)"
  ),
  WAIC = c(
    fit_single$WAIC,
    fit_change$WAIC,
    fit_triple$WAIC
  )
)

waic_available <- all(is.finite(comparison$WAIC))

if (waic_available) {
  comparison <- comparison |>
    arrange(WAIC) |>
    mutate(
      delta_WAIC = WAIC - min(WAIC),
      relative_likelihood = exp(-0.5 * delta_WAIC),
      WAIC_weight = relative_likelihood / sum(relative_likelihood),
      rank = row_number()
    )
} else {
  comparison <- comparison |>
    mutate(
      delta_WAIC = NA_real_,
      relative_likelihood = NA_real_,
      WAIC_weight = NA_real_,
      rank = NA_integer_
    )
}

knitr::kable(comparison, digits = 3)
write.csv(
  comparison,
  "output/model_comparison_WAIC.csv",
  row.names = FALSE
)
```

**How to read this:** lower WAIC is better. A very small difference means that the data do not clearly distinguish the competing demographic stories. The most complicated model should not be preferred merely because it can draw a more complicated curve.

```{r model-comparison-graphic, fig.width=9, fig.height=5.4}
if (waic_available) {
  p_waic <- comparison |>
    arrange(delta_WAIC) |>
    mutate(model = factor(model, levels = rev(model))) |>
    ggplot(aes(delta_WAIC, model)) +
    geom_col(width = 0.62) +
    geom_text(
      aes(label = sprintf("ΔWAIC %.2f", delta_WAIC)),
      hjust = -0.05,
      size = 4
    ) +
    labs(
      title = "Which demographic model is best supported?",
      subtitle = "Lower ΔWAIC means stronger predictive support",
      x = expression(Delta * "WAIC"),
      y = NULL,
      caption = "Kev-Carbon Bayesian model comparison"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank()
    ) +
    expand_limits(x = max(comparison$delta_WAIC) * 1.18 + 0.1)

  print(p_waic)
  ggsave(
    "output/01_model_comparison.png",
    p_waic,
    width = 9,
    height = 5.4,
    dpi = 320
  )
}
```

# 7. Posterior summaries and convergence

```{r posterior-diagnostics}
as_mcmc_list <- function(x) {
  if (inherits(x, "mcmc.list")) return(x)
  if (inherits(x, "mcmc")) return(coda::mcmc.list(x))
  if (is.list(x) && all(vapply(x, inherits, logical(1), "mcmc"))) {
    return(coda::mcmc.list(x))
  }
  stop("Posterior samples are not a coda mcmc or mcmc.list object.")
}

single_samples <- as_mcmc_list(fit_single$samples)
change_samples <- as_mcmc_list(fit_change$samples)
triple_samples <- as_mcmc_list(fit_triple$samples)

summarise_parameter <- function(samples, model_name) {
  mat <- as.matrix(samples)
  qs <- apply(mat, 2, quantile, probs = c(0.025, 0.5, 0.975))

  tibble(
    model = model_name,
    parameter = colnames(mat),
    mean = colMeans(mat),
    median = qs[2, ],
    lower_95 = qs[1, ],
    upper_95 = qs[3, ],
    effective_sample_size = as.numeric(coda::effectiveSize(samples))
  )
}

posterior_summary <- bind_rows(
  summarise_parameter(single_samples, "Single slope"),
  summarise_parameter(change_samples, "One change point"),
  summarise_parameter(triple_samples, "Two change points")
)

knitr::kable(posterior_summary, digits = 5)
write.csv(
  posterior_summary,
  "output/posterior_parameter_summary.csv",
  row.names = FALSE
)

rhat_table <- function(samples, model_name) {
  mat <- as.matrix(samples)

  if (length(samples) < 2) {
    return(tibble(
      model = model_name,
      parameter = colnames(mat),
      Rhat = NA_real_
    ))
  }

  gd <- tryCatch(
    coda::gelman.diag(
      samples,
      autoburnin = FALSE,
      multivariate = FALSE
    )$psrf[, 1],
    error = function(e) rep(NA_real_, ncol(mat))
  )

  tibble(
    model = model_name,
    parameter = colnames(mat),
    Rhat = as.numeric(gd)
  )
}

convergence <- bind_rows(
  rhat_table(single_samples, "Single slope"),
  rhat_table(change_samples, "One change point"),
  rhat_table(triple_samples, "Two change points")
)

knitr::kable(convergence, digits = 3)
write.csv(
  convergence,
  "output/convergence_Rhat.csv",
  row.names = FALSE
)
```

# 8. What does the new model actually say?

```{r plain-english}
triple_mat <- as.matrix(triple_samples)

q_r1 <- quantile(triple_mat[, "r1"], c(.025, .5, .975))
q_r2 <- quantile(triple_mat[, "r2"], c(.025, .5, .975))
q_r3 <- quantile(triple_mat[, "r3"], c(.025, .5, .975))

q_mu1 <- quantile(triple_mat[, "mu1"], c(.025, .5, .975))
q_mu2 <- quantile(triple_mat[, "mu2"], c(.025, .5, .975))

p1_growth <- mean(triple_mat[, "r1"] > 0)
p2_decline <- mean(triple_mat[, "r2"] < 0)
p3_growth <- mean(triple_mat[, "r3"] > 0)

p_decline_recovery <- mean(
  triple_mat[, "r2"] < 0 &
  triple_mat[, "r3"] > 0
)

p_accelerating_decline <- mean(
  triple_mat[, "r2"] < 0 &
  triple_mat[, "r3"] < triple_mat[, "r2"]
)

cat("## Plain-English interpretation\n\n")

cat(
  sprintf(
    "**First phase:** the posterior probability of growth is **%.1f%%**. ",
    100 * p1_growth
  ),
  sprintf(
    "The median rate is %.3f%% per year (95%% credible interval %.3f%% to %.3f%%).\n\n",
    100 * q_r1[2], 100 * q_r1[1], 100 * q_r1[3]
  )
)

cat(
  sprintf(
    "**First transition:** centred on **%.0f cal BP (about AD %.0f)**, ",
    q_mu1[2], bp_to_ad(q_mu1[2])
  ),
  sprintf(
    "with a 95%% credible interval of %.0f–%.0f cal BP.\n\n",
    q_mu1[1], q_mu1[3]
  )
)

cat(
  sprintf(
    "**Middle phase:** the posterior probability of decline is **%.1f%%**. ",
    100 * p2_decline
  ),
  sprintf(
    "The median rate is %.3f%% per year.\n\n",
    100 * q_r2[2]
  )
)

cat(
  sprintf(
    "**Second transition:** centred on **%.0f cal BP (about AD %.0f)**, ",
    q_mu2[2], bp_to_ad(q_mu2[2])
  ),
  sprintf(
    "with a 95%% credible interval of %.0f–%.0f cal BP.\n\n",
    q_mu2[1], q_mu2[3]
  )
)

cat(
  sprintf(
    "**Final phase:** the posterior probability of positive growth/recovery is **%.1f%%**. ",
    100 * p3_growth
  ),
  sprintf(
    "The median rate is %.3f%% per year.\n\n",
    100 * q_r3[2]
  )
)

cat(
  sprintf(
    "**Combined pattern:** P(decline followed by recovery) = **%.1f%%**; ",
    100 * p_decline_recovery
  ),
  sprintf(
    "P(decline followed by an even more negative rate) = **%.1f%%**.\n\n",
    100 * p_accelerating_decline
  )
)

plain_results <- tibble(
  quantity = c(
    "P phase 1 growth",
    "P phase 2 decline",
    "P phase 3 growth/recovery",
    "P decline then recovery",
    "P accelerating decline",
    "Median first transition cal BP",
    "Median second transition cal BP",
    "Median first transition cal AD",
    "Median second transition cal AD"
  ),
  value = c(
    p1_growth,
    p2_decline,
    p3_growth,
    p_decline_recovery,
    p_accelerating_decline,
    q_mu1[2],
    q_mu2[2],
    bp_to_ad(q_mu1[2]),
    bp_to_ad(q_mu2[2])
  )
)

write.csv(
  plain_results,
  "output/plain_english_results.csv",
  row.names = FALSE
)
```

> **Interpretive limit:** even a strongly supported decline does not by itself demonstrate its cause. The estimated timing must be compared with historical and archaeological evidence, while considering sampling intensity, preservation, mobility, research history and changes in depositional behaviour.

# 9. Graphic 1 — the demographic trajectories

```{r trajectory-graphic, fig.width=10, fig.height=6.2}
single_mat <- as.matrix(single_samples)
change_mat <- as.matrix(change_samples)

grid_bp <- seq(start_bp, end_bp, by = -5)

thin_draws <- function(mat, maximum = 350) {
  if (nrow(mat) <= maximum) return(mat)
  mat[round(seq(1, nrow(mat), length.out = maximum)), , drop = FALSE]
}

single_draws <- thin_draws(single_mat)
change_draws <- thin_draws(change_mat)
triple_draws <- thin_draws(triple_mat)

single_curves <- t(apply(single_draws, 1, function(z) {
  vapply(
    grid_bp,
    function(xx) nimbleCarbon::dExponentialGrowth(
      x = xx,
      a = start_bp,
      b = end_bp,
      r = z["r"],
      log = FALSE
    ),
    numeric(1)
  )
}))

change_curves <- t(apply(change_draws, 1, function(z) {
  vapply(
    grid_bp,
    function(xx) nimbleCarbon::dDoubleExponentialGrowth(
      x = xx,
      a = start_bp,
      b = end_bp,
      r1 = z["r_early"],
      r2 = z["r_late"],
      mu = z["mu"],
      log = FALSE
    ),
    numeric(1)
  )
}))

triple_curves <- t(apply(triple_draws, 1, function(z) {
  triple_density_vec(
    x = grid_bp,
    a = start_bp,
    b = end_bp,
    r1 = z["r1"],
    r2 = z["r2"],
    r3 = z["r3"],
    mu1 = z["mu1"],
    mu2 = z["mu2"]
  )
}))

curve_summary <- function(curves, model_name) {
  tibble(
    calBP = grid_bp,
    model = model_name,
    median = apply(curves, 2, median, na.rm = TRUE),
    lower = apply(curves, 2, quantile, .025, na.rm = TRUE),
    upper = apply(curves, 2, quantile, .975, na.rm = TRUE)
  )
}

curve_data <- bind_rows(
  curve_summary(single_curves, "Single slope"),
  curve_summary(change_curves, "One change point"),
  curve_summary(triple_curves, "Two change points")
) |>
  mutate(calendar_display_value = calendar_values(calBP))

write.csv(
  curve_data,
  "output/posterior_model_curves.csv",
  row.names = FALSE
)

p_curve <- ggplot(
  curve_data,
  aes(calendar_display_value, median, linetype = model)
) +
  geom_ribbon(
    aes(ymin = lower, ymax = upper, group = model),
    alpha = .13,
    linewidth = 0
  ) +
  geom_line(linewidth = 1.05) +
  calendar_scale +
  labs(
    title = "Three competing demographic histories",
    subtitle = "Posterior median trajectories with 95% credible intervals",
    x = calendar_axis_label,
    y = "Relative demographic intensity",
    linetype = "Model",
    caption = "The curves compare shape, not absolute population size."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

print(p_curve)

ggsave(
  "output/02_demographic_trajectories.png",
  p_curve,
  width = 10,
  height = 6.2,
  dpi = 320
)
```

# 10. Graphic 2 — when did the two transitions occur?

```{r transition-graphic, fig.width=9, fig.height=6.5}
transition_data <- tibble(
  `First transition` = triple_mat[, "mu1"],
  `Second transition` = triple_mat[, "mu2"]
) |>
  pivot_longer(
    everything(),
    names_to = "transition",
    values_to = "calBP"
  ) |>
  mutate(calendar_display_value = calendar_values(calBP))

p_transitions <- ggplot(
  transition_data,
  aes(calendar_display_value)
) +
  geom_density(
    fill = "grey75",
    alpha = .62,
    linewidth = .9
  ) +
  facet_wrap(
    ~ transition,
    ncol = 1,
    scales = "free_y"
  ) +
  calendar_scale +
  labs(
    title = "When did demographic change occur?",
    subtitle = "Posterior probability distributions for the two Kev-Carbon transition dates",
    x = calendar_axis_label,
    y = "Posterior density"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold")
  )

print(p_transitions)

ggsave(
  "output/03_transition_dates.png",
  p_transitions,
  width = 9,
  height = 6.5,
  dpi = 320
)
```

# 11. Graphic 3 — growth or decline in each phase?

```{r rates-graphic, fig.width=9, fig.height=7}
rates_data <- tibble(
  `Phase 1 — early` = 100 * triple_mat[, "r1"],
  `Phase 2 — middle` = 100 * triple_mat[, "r2"],
  `Phase 3 — late` = 100 * triple_mat[, "r3"]
) |>
  pivot_longer(
    everything(),
    names_to = "phase",
    values_to = "rate_pct"
  )

p_rates <- ggplot(rates_data, aes(rate_pct)) +
  geom_density(
    fill = "grey75",
    alpha = .62,
    linewidth = .9
  ) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    linewidth = .8
  ) +
  facet_wrap(
    ~ phase,
    ncol = 1,
    scales = "free_y"
  ) +
  labs(
    title = "Growth or decline in each demographic phase?",
    subtitle = "Values above zero indicate growth; values below zero indicate decline",
    x = "Estimated demographic rate (% per year)",
    y = "Posterior density"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold")
  )

print(p_rates)

ggsave(
  "output/04_phase_rates.png",
  p_rates,
  width = 9,
  height = 7,
  dpi = 320
)
```

# 12. Graphic 4 — the Kev-Carbon result at a glance

This final graphic is intended to be readable without knowing NIMBLE or Bayesian notation.

```{r result-at-glance, fig.width=10, fig.height=5.5}
phase_summary <- tibble(
  phase = factor(
    c("Phase 1", "Phase 2", "Phase 3"),
    levels = c("Phase 1", "Phase 2", "Phase 3")
  ),
  median_rate = 100 * c(q_r1[2], q_r2[2], q_r3[2]),
  lower = 100 * c(q_r1[1], q_r2[1], q_r3[1]),
  upper = 100 * c(q_r1[3], q_r2[3], q_r3[3]),
  plain = c(
    sprintf("%.0f%% probability of growth", 100 * p1_growth),
    sprintf("%.0f%% probability of decline", 100 * p2_decline),
    sprintf("%.0f%% probability of growth/recovery", 100 * p3_growth)
  )
)

p_at_glance <- ggplot(
  phase_summary,
  aes(phase, median_rate)
) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    linewidth = .7
  ) +
  geom_errorbar(
    aes(ymin = lower, ymax = upper),
    width = .12,
    linewidth = 1
  ) +
  geom_point(size = 3.6) +
  geom_text(
    aes(label = plain),
    vjust = -1.25,
    size = 4
  ) +
  labs(
    title = "Kev-Carbon result at a glance",
    subtitle = "Posterior median demographic rates and 95% credible intervals",
    x = NULL,
    y = "Demographic rate (% per year)",
    caption = paste0(
      "Estimated transitions: ",
      round(q_mu1[2]), " cal BP and ",
      round(q_mu2[2]), " cal BP"
    )
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 17),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )

print(p_at_glance)

ggsave(
  "output/05_result_at_a_glance.png",
  p_at_glance,
  width = 10,
  height = 5.5,
  dpi = 320
)
```

# 13. Automatic conclusion

```{r automatic-conclusion, results='asis'}
if (waic_available) {

  winner <- comparison$model[which.min(comparison$WAIC)]
  best_delta_other <- sort(comparison$delta_WAIC)[2]

  cat("### Best-supported model\n\n")
  cat("The lowest WAIC is obtained by **", winner, "**.\n\n", sep = "")

  if (best_delta_other < 2) {
    cat(
      "**Caution:** the leading model is not decisively separated from its nearest competitor ",
      "(ΔWAIC < 2). The data therefore support more than one plausible demographic description.\n\n"
    )
  }

  if (winner == "Two change points (Kev-Carbon)") {
    cat(
      "In plain English, the radiocarbon evidence is better described by **three demographic phases** ",
      "than by either a single continuous trend or one abrupt change. The estimated transition dates ",
      "and their credible intervals should now be compared with independent archaeological and historical evidence.\n\n"
    )
  }

  if (winner == "One change point") {
    cat(
      "In plain English, the data support **one major change in demographic rate**. ",
      "The additional complexity of a second transition is not sufficiently supported by predictive fit.\n\n"
    )
  }

  if (winner == "Single slope") {
    cat(
      "In plain English, these data do not require a distinct transition to explain the broad trajectory. ",
      "This is not evidence that historical events had no demographic effect; it means the radiocarbon evidence ",
      "does not resolve them as a separate change in the fitted temporal density.\n\n"
    )
  }
}
```

# 14. What Kev-Carbon adds

The scientific contribution is not simply a prettier curve. The extension adds a formally estimated second demographic transition while retaining the Bayesian radiocarbon modelling framework.

The key questions now become:

1. Does the additional transition improve predictive performance?
2. Can simulation recover the true transition dates and rates?
3. How uncertain are the inferred transition dates?
4. Is a decline followed by recovery more probable than continued or accelerating decline?
5. Do these results remain stable under alternative analytical windows, bin widths and priors?

Those questions form the next validation stage before treating the model as a general archaeological method.

# 15. Session information

```{r session-info}
sessionInfo()
```


# 16. How to run this version

1. Put this `.Rmd` in the same project folder as `AusData.xlsx`.
2. Make sure Rtools45 is installed and `pkgbuild::check_build_tools(debug = TRUE)` reports that the build tools are ready.
3. Open the `.Rmd` in RStudio.
4. Leave `run_mode: "test"` for the first run.
5. Knit to HTML.
6. Check that the `output/` folder contains:
   - `fit_single_slope.rds`
   - `fit_change_point.rds`
   - `fit_two_change_points.rds`
   - `model_comparison_WAIC.csv`
   - `posterior_parameter_summary.csv`
   - `plain_english_results.csv`
   - `01_model_comparison.png`
   - `02_demographic_trajectories.png`
   - `03_transition_dates.png`
   - `04_phase_rates.png`
   - `05_result_at_a_glance.png`
7. Only after the test run succeeds, change the YAML to `run_mode: "medium"`.

**What to send back if it fails:** copy the first error beginning with `NIMBLE ERROR`, plus roughly 20 lines above it. That will identify whether the problem is compilation, initial values, the custom distribution, WAIC extraction or plotting.
