---
title: "AusDecline: Integrated Bayesian Master Report"
author: "Kevan Edinborough"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    theme: flatly
    highlight: tango
---

```{r master-setup, include=FALSE, cache=TRUE}
find_project_root <- function(start = getwd()) {
  candidates <- c(Sys.getenv("AUSDECLINE_PROJECT_ROOT", unset = ""), start)
  for (candidate in candidates[nzchar(candidates)]) {
    here <- normalizePath(candidate, winslash = "/", mustWork = TRUE)
    repeat {
      if (file.exists(file.path(here, "config.yml")) &&
          file.exists(file.path(here, "R", "00_setup.R"))) return(here)
      parent <- dirname(here)
      if (identical(parent, here)) break
      here <- parent
    }
  }
  stop("Could not locate the AusDecline project root.")
}
PROJECT_ROOT <- find_project_root()
setwd(PROJECT_ROOT)
knitr::opts_knit$set(root.dir = PROJECT_ROOT)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 5.8, dpi = 160)
source(file.path(PROJECT_ROOT, "R", "00_setup.R"))
source(file.path(PROJECT_ROOT, "R", "10_model_explanations.R"))
source(file.path(PROJECT_ROOT, "R", "01_import_clean.R"))
source(file.path(PROJECT_ROOT, "R", "02_calibrate_bin.R"))
source(file.path(PROJECT_ROOT, "R", "03_define_models.R"))
source(file.path(PROJECT_ROOT, "R", "04_fit_models.R"))
source(file.path(PROJECT_ROOT, "R", "05_compare_waic.R"))
source(file.path(PROJECT_ROOT, "R", "06_diagnostics.R"))
source(file.path(PROJECT_ROOT, "R", "07_posterior_curves.R"))

fmt_year <- function(bp) {
  if (calendar_display == "calad") {
    y <- bp_to_ad(bp)
    ifelse(y <= 0, paste0(abs(round(y)), " BC"), paste0("AD ", round(y)))
  } else paste0(round(bp), " cal BP")
}

best_key <- if (waic_available) comparison$model_key[1] else NA_character_
best_model <- if (waic_available) comparison$model[1] else "No valid WAIC winner"
best_weight <- if (waic_available) comparison$WAIC_weight[1] else NA_real_
best_delta <- if (waic_available && nrow(comparison) > 1) comparison$delta_WAIC[2] else NA_real_

max_rhat <- if (nrow(convergence) && any(is.finite(convergence$Rhat))) max(convergence$Rhat, na.rm=TRUE) else NA_real_
min_ess <- if (nrow(posterior_summary) && any(is.finite(posterior_summary$effective_sample_size))) min(posterior_summary$effective_sample_size, na.rm=TRUE) else NA_real_

plain_verdict <- if (!waic_available) {
  "All enabled models fitted, but a valid WAIC ranking was not available; no preferred model is declared."
} else switch(best_key,
  single = "The radiocarbon-derived trajectory is most economically described by one broad proportional rate across the selected interval.",
  change = "The evidence favours different early and late rates separated by an estimated abrupt transition date.",
  logistic = "The evidence favours a gradual curved transition centred on an estimated inflection date rather than an instantaneous break.",
  "No interpretation available."
)
```

<div style="background:#eef6f8;border-left:8px solid #176B87;padding:18px 24px;border-radius:8px;margin-bottom:24px;">
<h2 style="margin-top:0;color:#183B4E;">Executive summary</h2>
<p><strong>Analysis window:</strong> `r fmt_year(start_bp)` to `r fmt_year(end_bp)` (`r start_bp`–`r end_bp` cal BP).</p>
<p><strong>Dataset:</strong> `r nrow(AUD)` valid determinations from `r dplyr::n_distinct(AUD$site_id)` sites, represented by `r nrow(binned)` independent site-time bins.</p>
<p><strong>Best-supported model:</strong> `r best_model`.</p>
<p>`r plain_verdict`</p>
<p><strong>Reliability:</strong> maximum R-hat = `r ifelse(is.finite(max_rhat), sprintf('%.3f', max_rhat), 'unavailable')`; minimum effective sample size = `r ifelse(is.finite(min_ess), round(min_ess), 'unavailable')`.</p>
</div>

# Research aim

This project tests which of three temporal shapes best describes the distribution of calibrated Australian radiocarbon evidence across the configured interval:

1. **Single slope:** one proportional rate throughout.
2. **Change point:** one early rate and one late rate separated by an estimated abrupt transition.
3. **Logistic transition:** a gradual curved transition centred on an estimated inflection date.

The analysis evaluates temporal shape. It does not, by itself, prove population size or establish European colonisation as the sole cause of any decline.

# Plain-English guide to the models

## Single-slope model

`r model_plain_english$single`

## Change-point model

`r model_plain_english$change`

## Logistic-transition model

`r model_plain_english$logistic`

# Data and configuration

## Active global settings

```{r configuration-table}
knitr::kable(configuration, align = c("l", "l"))
```

## Dataset audit

```{r audit-table}
knitr::kable(audit, digits = 1)
```

The workflow imports `LabID`, `SiteID`, conventional radiocarbon age and laboratory error. Internal calibration identifiers are made unique, preserving repeated laboratory identifiers in the original data.

# Calibration and temporal independence

Dates are calibrated with **`r config$calibration_curve`** using `rcarbon::calibrate()`. Calibration normalisation is set to **`r isTRUE(config$calibration_normalised)`**. Same-site determinations close in conventional radiocarbon age are grouped with a bin width of **`r config$site_bin_width_years` years**. One median calibrated calendar estimate is retained per site-time bin.

```{r binned-distribution}
print(
  ggplot(binned_plot, aes(calendar_display_value)) +
    geom_histogram(binwidth = 50, boundary = 0, fill = "#176B87", colour = "white") +
    calendar_scale +
    labs(title = "Independent site-time bins used in the models",
         subtitle = paste(nrow(binned), "bins representing", sum(binned$dates_in_bin), "determinations"),
         x = calendar_axis_label, y = "Site-time bins") +
    theme_minimal(base_size = 13)
)
```

# Statistical methods

## Shared likelihood framework

The binned calendar estimates are treated as observations from demographic date-density distributions supplied by `nimbleCarbon`. All enabled models use the same data, analysis interval, calibration and binning, so the WAIC comparison is conditional on the same preprocessing.

## Model 1: single slope

`r model_method_details$single`

The prior is \(r \sim \mathrm{Normal}(0, 0.02)\) on the rate scale.

## Model 2: change point

`r model_method_details$change`

The early and late rates each receive `Normal(0, 0.02)` priors. The change point is uniform within an interior interval that excludes the temporal boundaries.

## Model 3: logistic transition

`r model_method_details$logistic`

The logistic rate receives a `Normal(0, 0.02)` prior. The inflection date is uniform within the same interior interval used for the change-point model.

## MCMC specification

The active run used **`r mcmc_cfg$nchains` chains**, **`r mcmc_cfg$niter` iterations per chain**, **`r mcmc_cfg$nburnin` burn-in iterations**, and thinning of **`r mcmc_cfg$thin`**. It retained approximately **`r (mcmc_cfg$niter-mcmc_cfg$nburnin)%/%mcmc_cfg$thin` samples per chain**. Models were fitted on **`r parallel_cores` processor(s)**.

# Comparative results

## WAIC model comparison

```{r waic-table}
knitr::kable(comparison, digits = 3)
```

```{r waic-plot}
if (waic_available) {
  wp <- comparison |>
    dplyr::mutate(model = factor(model, levels = rev(model)))
  print(
    ggplot(wp, aes(WAIC_weight, model)) +
      geom_col(fill = "#176B87", width = 0.65) +
      geom_text(aes(label = sprintf("%.1f%%", 100*WAIC_weight)), hjust = -0.1, fontface = "bold") +
      scale_x_continuous(limits = c(0, 1.12), labels = function(x) paste0(round(100*x), "%")) +
      labs(title = "Relative support for candidate decline models", x = "WAIC weight", y = NULL) +
      theme_minimal(base_size = 13) + theme(panel.grid.major.y = element_blank())
  )
} else cat("WAIC weights are unavailable; no model-support graph is produced.")
```

A lower WAIC indicates better expected out-of-sample predictive performance among the candidate models. WAIC weights express relative support within this candidate set; they are not probabilities that a model is historically true.

## Posterior parameter estimates

```{r posterior-table}
knitr::kable(posterior_summary, digits = 5)
```

## Direct temporal questions

```{r slope-question-table}
knitr::kable(slope_questions, digits = 4)
```

# Convergence and reliability

```{r convergence-table}
knitr::kable(convergence, digits = 3)
```

R-hat values close to 1 indicate agreement between chains. Effective sample size measures the amount of independent posterior information. Results with high R-hat or very low effective sample size should be rerun in medium or long mode before substantive interpretation.

# Posterior decline shapes

```{r posterior-curves}
print(p_curve)
```

```{r posterior-slopes}
print(p_slope)
```

The curve figure compares fitted temporal shapes after normalisation. The slope figure shows change as chronology moves toward the present; values below zero represent decline toward the present.

# Model-specific results in plain English

## Single slope

`r model_result_text("single", matrices_by_model, calendar_display, bp_to_ad)`

## Change point

`r model_result_text("change", matrices_by_model, calendar_display, bp_to_ad)`

## Logistic transition

`r model_result_text("logistic", matrices_by_model, calendar_display, bp_to_ad)`

# Integrated model verdict

```{r verdict, results='asis'}
source(file.path(PROJECT_ROOT, "R", "08_outputs_interpretation.R"))
```

# Archaeological interpretation

A change-point or logistic transition whose uncertainty overlaps historically documented phases of European occupation may be **consistent with** colonial disruption. Relevant mechanisms could include disease, frontier violence, dispossession, restrictions on mobility, altered food systems, forced relocation and changes in archaeological deposition. However, temporal overlap alone does not demonstrate causation.

If the single-slope model is favoured, the radiocarbon record does not require a distinct colonial-period break. That result would not imply that colonisation had no effect; it would mean that the chosen data and models do not resolve a separate rate change.

# Sensitivity analyses required before publication

The principal results should be repeated across:

- analysis windows beginning at 1000, 2000, 3000 and 4000 cal BP;
- alternative site-bin widths;
- plausible calibration choices and normalisation settings;
- medium and long MCMC runs;
- alternative random seeds.

A transition date that remains stable across these choices is more credible than one that shifts substantially with the analysis window.

# Limitations

Radiocarbon frequency is an archaeological proxy, not a direct census. It is affected by sampling intensity, preservation, research history, site destruction, mobility, depositional practices, calibration-curve structure, chronological uncertainty, regional aggregation and the chosen model family. The present analysis uses one representative calendar estimate per site-time bin; future hierarchical models should propagate full calibrated-date uncertainty directly.

# Reproducibility record

```{r output-files}
files <- list.files(OUTPUT_DIR, recursive = TRUE, full.names = FALSE)
knitr::kable(tibble::tibble(output_file = files))
```

# Session information

```{r session-info}
sessionInfo()
```
