---
title: "Single-slope model: detailed methods and results"
author: "Kevan Edinborough"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: readable
params:
  model_key: "single"
  report_title: "Single-slope model: detailed methods and results"
---

# Research question

```{r setup, message=FALSE, warning=FALSE}
find_project_root <- function(start = getwd()) {
  here <- normalizePath(start, winslash = "/", mustWork = TRUE)
  repeat {
    if (file.exists(file.path(here, "config.yml")) && file.exists(file.path(here, "R", "00_setup.R"))) return(here)
    parent <- dirname(here)
    if (identical(parent, here)) stop("Could not locate project root.")
    here <- parent
  }
}
PROJECT_ROOT <- find_project_root()
setwd(PROJECT_ROOT)
knitr::opts_knit$set(root.dir = PROJECT_ROOT)
knitr::opts_chunk$set(echo=TRUE, message=TRUE, warning=TRUE, fig.width=9, fig.height=5.5)
source(file.path(PROJECT_ROOT, "R", "00_setup.R"))
source(file.path(PROJECT_ROOT, "R", "10_model_explanations.R"))
model_key <- params$model_key
cat(model_plain_english[[model_key]])
```

# Configuration

```{r configuration}
knitr::kable(configuration)
```

# Data preparation

```{r import}
source(file.path(PROJECT_ROOT, "R", "01_import_clean.R"))
knitr::kable(audit, digits=1)
```

```{r calibrate-bin, cache=TRUE}
source(file.path(PROJECT_ROOT, "R", "02_calibrate_bin.R"))
```

The dates are calibrated using the curve specified in the active configuration. Determinations from the same site and similar radiocarbon age are grouped into site-time bins to reduce the influence of oversampled contexts.

# Detailed model method

```{r method}
source(file.path(PROJECT_ROOT, "R", "03_define_models.R"))
cat(model_method_details[[model_key]])
```

The model is estimated by MCMC in NIMBLE using the chain count, iterations, burn-in and thinning specified in the active configuration. WAIC may be printed for descriptive inspection, but a one-model run cannot establish relative model preference.

# Model fitting

```{r fit-model, cache=TRUE}
source(file.path(PROJECT_ROOT, "R", "04_fit_models.R"))
```

# Posterior results and diagnostics

```{r diagnostics}
source(file.path(PROJECT_ROOT, "R", "06_diagnostics.R"))
knitr::kable(posterior_summary, digits=5)
knitr::kable(convergence, digits=3)
knitr::kable(slope_questions, digits=4)
```

```{r curves}
source(file.path(PROJECT_ROOT, "R", "07_posterior_curves.R"))
print(p_curve)
print(p_slope)
```

# Plain-English result

```{r interpretation, results='asis'}
cat("**Model:** ", model_labels[[model_key]], "  \n\n", sep="")
cat(model_result_text(model_key, matrices_by_model, calendar_display, bp_to_ad), "  \n\n")
cat("This result describes the temporal shape supported within this model. It does not by itself prove population size, colonial causation, or superiority over the other candidate models. Use the all-model comparison report for that comparison.")
```

# Methodological limitations

Radiocarbon frequency is an archaeological proxy affected by sampling intensity, preservation, research history, site destruction, mobility, depositional practice, calibration-curve structure and the selected time window. Colonial interpretation therefore requires comparison with independent historical evidence and sensitivity analyses.

# Session information

```{r session-info}
sessionInfo()
```
