---
title: "Bayesian comparison of three Australian radiocarbon decline models"
author: "Kevan Edinborough"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: readable
---

# Aim

This report compares three demographic models over the interval set in `config.yml`.

1. **Single-slope model:** one proportional rate across the complete interval.
2. **Change-point model:** an early rate, a later rate and an estimated abrupt change date.
3. **Smooth logistic transition:** a gradual S-shaped change centred on an estimated inflection date, allowing decline to bend and stabilise.

The comparison tests whether the radiocarbon-derived trajectory is best represented by one broad rate, an abrupt change in rate, or a smooth transition. It does not by itself prove that European colonisation caused the inferred change.

# 1. Configuration

```{r setup, message=FALSE, warning=FALSE}
# Locate the project root robustly whether Knit is clicked from the project
# folder, the reports folder, or run through rmarkdown::render().
find_project_root <- function(start = getwd()) {
  here <- normalizePath(start, winslash = "/", mustWork = TRUE)
  repeat {
    if (file.exists(file.path(here, "config.yml")) &&
        file.exists(file.path(here, "R", "00_setup.R"))) {
      return(here)
    }
    parent <- dirname(here)
    if (identical(parent, here)) {
      stop("Could not locate the project root containing config.yml and R/00_setup.R.")
    }
    here <- parent
  }
}

PROJECT_ROOT <- find_project_root()
setwd(PROJECT_ROOT)
knitr::opts_knit$set(root.dir = PROJECT_ROOT)
knitr::opts_chunk$set(echo=TRUE, message=TRUE, warning=TRUE, fig.width=9, fig.height=5.5)
source(file.path(PROJECT_ROOT, "R", "00_setup.R"))
knitr::kable(configuration)
cat("Resolved window:", start_bp, "to", end_bp, "cal BP (", bp_to_ad(start_bp), "to", bp_to_ad(end_bp), "AD).")
```

# 2. Import and audit

```{r import}
source(file.path(PROJECT_ROOT, "R", "01_import_clean.R"))
knitr::kable(audit, digits=1)
```

# 3. Calibration and site-time bins

```{r calibrate-bin, cache=TRUE}
source(file.path(PROJECT_ROOT, "R", "02_calibrate_bin.R"))
```

The same binned calendar dates are used by all candidate models.

# 4. Candidate models

```{r define-models}
source(file.path(PROJECT_ROOT, "R", "03_define_models.R"))
```

The **single-slope model** asks whether one continuous proportional trend is sufficient. The **change-point model** asks whether the rate changes abruptly at an unknown date. The **logistic model** asks whether the trajectory bends gradually around an inflection date and then stabilises.

# 5. Fit models

```{r fit-models, cache=TRUE}
source(file.path(PROJECT_ROOT, "R", "04_fit_models.R"))
```

# 6. WAIC comparison

```{r waic}
source(file.path(PROJECT_ROOT, "R", "05_compare_waic.R"))
knitr::kable(comparison, digits=3)
```

WAIC ranking is reported only when valid finite values are extracted for every enabled model.

# 7. Posterior diagnostics

```{r diagnostics}
source(file.path(PROJECT_ROOT, "R", "06_diagnostics.R"))
knitr::kable(posterior_summary, digits=5)
knitr::kable(convergence, digits=3)
knitr::kable(slope_questions, digits=4)
```

# 8. Posterior model curves

```{r curves}
source(file.path(PROJECT_ROOT, "R", "07_posterior_curves.R"))
p_curve
p_slope
```

# 9. Model verdict

```{r outputs, results='asis'}
source(file.path(PROJECT_ROOT, "R", "08_outputs_interpretation.R"))
```

# 10. Session information

```{r session-info}
sessionInfo()
```
